---
title: "Longitudinal Final Report"
date: "2025-04-26"
output: pdf_document
header-includes:
  - \usepackage{booktabs}
  - \usepackage{multirow}
  - \usepackage{rotating}
  - \usepackage{array}
---
```{r}
library(haven)
library(dplyr)
library(tidyr)
library(MASS)
library(lme4)
library(nlme)
library(performance)
library(geepack)
library(ggplot2)
# library(afex) # will mask lmer

df <- read.csv("/Users/kangyiyuan/Desktop/2480 Final Project/data_final.csv")

head(df)

all_vars <- df %>% dplyr::select(-PID, -time) %>% names()

# Check time-variant and tme-invariant
is_time_variant <- function(var) {
  tmp <- df %>% dplyr::select(PID, time, all_of(var))
  var_name <- var
  var_change <- tmp %>%
    group_by(PID) %>%
    summarize(var_sd = sd(.data[[var_name]], na.rm = TRUE)) %>%
    pull(var_sd)
  
  mean(var_change > 0, na.rm = TRUE) > 0.08 # over 0.08 will be time variant
}

time_variant_vars <- all_vars[sapply(all_vars, is_time_variant)]
time_invariant_vars <- setdiff(all_vars, time_variant_vars)

time_variant_vars
time_invariant_vars

# Variable modification:
# Binary change
# Smoke treated as binary
df <- df %>%
  mutate(smoke_ever_clean = case_when(
    smoke_ever %in% c(1,0) ~ 1,   # Yes
    smoke_ever == 5  ~ 0,   # never
  ))

# new variable to indicate whether the observation is a current smoker or not
df <- df %>%
  mutate(current_smoker = case_when (
    smoke_ever == 0 ~ 1,
    smoke_ever %in% c(1, 5) ~ 0
  ))

# race change
problematic_pid <- df %>%
  group_by(PID) %>%
  summarise(n_race = n_distinct(race)) %>%
  filter(n_race > 1) %>%
  pull(PID)

# delete those PID
df <- df %>%
  filter(!(PID %in% problematic_pid))

# race change, merge race number 3-7 into the other category
df <- df %>% 
  mutate(race_clean = case_when(
    race == 1 ~ 1,
    race == 2 ~ 2,
    race %in% c(3, 4, 5, 7) ~ 3
  ))

# convert times to wave
df <- df %>% 
  mutate(wave = case_when(
    time == 7 ~ 0,
    time == 9 ~ 1,
    time == 11 ~ 2,
    time == 13 ~ 3
    ))
```

# data preparation
```{r}
exposure <- "discrimination"
main_out <- "mental_worry"
level_2 <- "race_clean"

df$race_clean <- as.factor(df$race_clean)

sub_vars <- c("smoke_ever_clean",
             "current_smoker",
             "marijuana_bi",
             "cocaine", 
             "dietpill", 
             "amphet", 
             "barbit", 
             "tranquil", 
             "steroids",
             "chronic_con",
             "bmi"
              )

sub_vars_d <- c("discrimination",
                "smoke_ever_clean",
             "current_smoker",
             "marijuana_bi",
             "cocaine", 
             "dietpill", 
             "amphet", 
             "barbit", 
             "tranquil", 
             "steroids",
             "chronic_con",
             "bmi"
              )

main_cov <- "discrimination"

df$race_labeled <- factor(df$race,
  levels = c(1, 2, 3, 4, 5, 7),
  labels = c(
    "1 White",
    "2 Black, AA or Negro",
    "3 American Indian or Alaska Native",
    "4 Asian",
    "5 Native Hawaiian or Pacific Islander",
    "7 Some other race"
  )
)

df$race_clean_labeled <- factor(df$race_clean,
  levels = c(1, 2, 3),
  labels = c(
    "1 White",
    "2 Black, AA or Negro",
    "3 Other"
  )
)
```

## factorize variables
```{r, include=FALSE}
# PID, time, discri, mari_occ, bmi, mental_worry, wave
sapply(df[, -c(1, 2, 3, 6, 9, 15, 20)], as.factor)
```


# OLS.

```{r}
library(broom)
library(purrr)

# identify different kinds of outcome
# categorical or numerical

# OLS with wave
run_ols_with_aic <- function(sub_var) {
  fm <- paste(main_out, "~ wave  +", sub_var)
  model <- lm(as.formula(fm), data = df)
  
  coef_info <- tidy(model) %>% filter(term == sub_var)
  
  tibble(
    substance = sub_var,
    estimate = coef_info$estimate,
    std_error = coef_info$std.error,
    p_value = coef_info$p.value,
    AIC = AIC(model),
    BIC = BIC(model)
  )
}

ols_results_with_aic <- map_dfr(sub_vars, run_ols_with_aic)

model_comparisons <- ols_results_with_aic %>%
  mutate(sig = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01  ~ "**",
    p_value < 0.05  ~ "*",
    p_value < 0.1   ~ ".",
    TRUE ~ ""
  ))

print(model_comparisons)
```

# General OLS only with wave
```{r}
worry_ols <- lm(mental_worry ~ wave, data = df)
summary(worry_ols)
cor(df$mental_worry,df$wave)
cov2cor(vcov(worry_ols))
AIC(worry_ols)
BIC(worry_ols)
```

```{r}
# individual growth plot (select 12 random individual)
set.seed(100)
race1_ids <- df %>%
  filter(race_clean_labeled == "1 White") %>%
  distinct(PID) %>%
  sample_n(5) %>%
  pull(PID)

race2_ids <- df %>%
  filter(race_clean_labeled == "2 Black, AA or Negro") %>%
  distinct(PID) %>%
  sample_n(4) %>%
  pull(PID)

race3_ids <- df %>%
  filter(race_clean_labeled == "3 Other") %>%
  distinct(PID) %>%
  sample_n(3) %>%
  pull(PID)
ids <- c(race1_ids, race2_ids, race3_ids)

df %>%
  filter(PID %in% ids) %>%
  # plot the empirical plots
  ggplot(mapping = aes(x = wave, y = mental_worry)) +
  geom_point() +
  geom_line() +
  facet_wrap(~PID) +
  labs(x = "Wave", y = "Worry Score",
       title = "The Empirical Growth Plots for 12 Participants")
df %>%
   filter(PID %in% ids) %>%
# plot the empirical plots
  ggplot(mapping = aes(x = wave, y = mental_worry)) + geom_point() + facet_wrap("PID") +
  geom_smooth(method = "lm", se = T, color = "blue") +
  labs(x = "Wave", y = "Worry Score",
       title = "The Empirical Growth Plots for 12 Participants")

#
individual_ols <- df %>%
  group_by(PID) %>%
  do(model = lm(mental_worry ~ wave, data = .))

interceptols <- slopeols <- NULL

for(i in 1:nrow(individual_ols)){
  interceptols[i] <- individual_ols[[2]][[i]][["coefficients"]][1]
  slopeols[i] <- individual_ols[[2]][[i]][["coefficients"]][2]
}

summary(interceptols)[4]
summary(slopeols)[4]

cor(interceptols, slopeols)

var(interceptols)
var(slopeols)

# report in table

library(ggplot2)

ggplot(df, mapping = aes(x = wave, y = mental_worry)) +
  geom_smooth(aes(group = as.factor(PID)), method = "lm", se = F, color =
                "black") + geom_smooth(method = "lm", se = F, color = "red") + 
  labs(x = "Wave", y = "Worry Score",
       title = "The Individual OLS Plots with Mean Trajectory.")

df %>%
  filter(PID %in% ids) %>%
  ggplot(mapping = aes(x = wave, y = mental_worry)) +
  geom_smooth(aes(group = as.factor(PID)), method = "lm", se = F, color =
                "black") + geom_smooth(method = "lm", se = F, color = "red") + 
  labs(x = "Wave", y = "Worry Score",
       title = "The Individual OLS of the 12 Plots with Mean Trajectory.")
```
On average, worry score remains stable over time.

# Panel plot

```{r}
library(ggplot2)
ggplot(df, aes(x = wave, y = mental_worry, group = PID)) +
  geom_line(alpha = 0.2) +
  stat_smooth(aes(group = 1), method = "lm", se = TRUE, color = "blue") +
  theme_minimal() +
  facet_wrap(~ race_clean_labeled)
```
```{r}
# OLS plots by race
df %>%
   filter(PID %in% ids) %>%
ggplot(aes(x = wave, y = mental_worry, group = PID, color = factor(PID))) +
  geom_line(linewidth = 0.5) +
  geom_point() +
  theme_minimal() +
  facet_wrap(~ race_clean_labeled) +
  stat_smooth(data = df, aes(group = race_clean_labeled), method = "lm", 
              color = "black", se = FALSE) +
  labs(color = "Participant ID", x = "Wave", y = "Worry Score", 
       title = "The Empirical Plots by Race") +
  scale_x_continuous(breaks = 0:3)
```

# lmer: model building

```{r}
library(lmerTest)
#ols_currentsmoker <-lm(mental_worry ~ wave + current_smoker, data = df_byid)
#summary(ols_currentsmoker)

# model.smoke.ever <- lme(mental_worry ~ wave + discrimination + smoke_ever_clean, data = df, random= ~ wave|PID, method="ML")
# summary(model.smoke.ever)

# # unconditional mean model
# model.a <- lmer(mental_worry ~ 1 + (1|PID), data = df, REML = FALSE)
# summary(model.a)
# 
# # uncon growth by race: level-2 variability small
# model.d <- lmer(mental_worry ~ wave + (wave|race_clean_labeled), data = df, REML = FALSE)
# summary(model.d)

# unconditional growth model
model.wave <- lmer(mental_worry ~ wave + (wave|PID), data = df, REML = FALSE)
summary(model.wave)
icc(model.wave)

# + discrimination
model.dis <- lmer(mental_worry ~ wave + discrimination + 
                    (wave|PID), data = df, REML = FALSE)
summary(model.dis)
icc(model.dis)

# + mari_bi 
model.dis.mari <- lmer(mental_worry ~ wave + discrimination + 
                         marijuana_bi + (wave|PID),
                       data = df, REML = FALSE)
summary(model.dis.mari)
icc(model.dis.mari)

# + ever smoke
model.dis.mari.esmoke <- lmer(mental_worry ~ wave + discrimination + 
                                marijuana_bi + smoke_ever_clean + (wave|PID),
                       data = df, REML = FALSE)
summary(model.dis.mari.esmoke)
icc(model.dis.mari.esmoke)

# + race
model.dis.mari.esmoke.race <- lmer(mental_worry ~ wave + discrimination +
                                     marijuana_bi + smoke_ever_clean + 
                                     (wave|PID) + race_clean_labeled,
                       data = df, REML = FALSE)
summary(model.dis.mari.esmoke.race)
icc(model.dis.mari.esmoke.race)

# Discussion
first_modelpoi <- glm(formula = mental_worry ~ wave, family = poisson(link = "log"), data = df)
summary(first_modelpoi)
final_modelpoi <- glm(formula = mental_worry ~ wave + discrimination + marijuana_bi + smoke_ever_clean + race_clean_labeled, family = poisson(link = "log"), data = df)
summary(final_modelpoi)

first_model_quasi <- glm(formula = mental_worry ~ wave, family = quasipoisson(), data = df)
summary(first_model_quasi)
final_model_quasi <- glm(formula = mental_worry ~ wave + discrimination + marijuana_bi + smoke_ever_clean + race_clean_labeled, family = quasipoisson(), data = df)
summary(final_model_quasi)
```

```{r}
model_list <- list(model.wave, model.dis, model.dis.mari, model.dis.mari.esmoke, 
                   model.dis.mari.esmoke.race)
aic_tab <- function(model){
  return(summary(model)$AICtab)
}

models_aic_tab <- sapply(model_list, aic_tab)
colnames(models_aic_tab) <- c("model.wave", "model.dis", "model.dis.mari", 
                              "model.dis.mari.esmoke","model.dis.mari.esmoke.race")
models_aic_tab
```

```{r}
# model.wave icc
icc.b <- icc(model.wave)$ICC_adjusted
icc_total.b <- sum(as.data.frame(VarCorr(model.wave),comp="Variance")$vcov[-3]) 
icc_intercept.b <- as.data.frame(VarCorr(model.wave),comp="Variance")$vcov[1]/icc_total.b
icc_wave.b <- as.data.frame(VarCorr(model.wave),comp="Variance")$vcov[2]/icc_total.b
icc_residual.b <- as.data.frame(VarCorr(model.wave),comp="Variance")$vcov[4]/icc_total.b

# model.dis icc
icc.dis <- icc(model.dis)$ICC_adjusted
icc_total.dis <- sum(as.data.frame(VarCorr(model.dis),comp="Variance")$vcov[-3])
icc_intercept.dis <- as.data.frame(VarCorr(model.dis),comp="Variance")$vcov[1]/icc_total.dis
icc_wave.dis <- as.data.frame(VarCorr(model.dis),comp="Variance")$vcov[2]/icc_total.dis
icc_residual.dis <- as.data.frame(VarCorr(model.dis),comp="Variance")$vcov[4]/icc_total.dis

# model.dis.mari
icc.dis.mari <- icc(model.dis.mari)$ICC_adjusted
icc_total.dismari <- sum(as.data.frame(VarCorr(model.dis.mari),
                                       comp="Variance")$vcov[-3])
icc_intercept.dismari <- as.data.frame(VarCorr(model.dis.mari),
                                       comp="Variance")$vcov[1]/icc_total.dismari
icc_wave.dismari <- as.data.frame(VarCorr(model.dis.mari),
                                  comp="Variance")$vcov[2]/icc_total.dismari
icc_residual.dismari <- as.data.frame(VarCorr(model.dis.mari),
                                      comp="Variance")$vcov[4]/icc_total.dismari

# model.dis.mari.esmoke
icc.dis.mari.esmoke <- icc(model.dis.mari.esmoke)$ICC_adjusted
icc_total.dis.mari.esmoke <- sum(as.data.frame(VarCorr(model.dis.mari.esmoke),
                                               comp="Variance")$vcov[-3])
icc_intercept.dis.mari.esmoke <- as.data.frame(VarCorr(model.dis.mari.esmoke),
                                               comp="Variance")$vcov[1]/icc_total.dis.mari.esmoke
icc_wave.dis.mari.esmoke <- as.data.frame(VarCorr(model.dis.mari.esmoke),
                                          comp="Variance")$vcov[2]/icc_total.dis.mari.esmoke
icc_residual.dis.mari.esmoke <- as.data.frame(VarCorr(model.dis.mari.esmoke),
                                              comp="Variance")$vcov[4]/icc_total.dis.mari.esmoke

# model.dis.mari.esmoke + race
icc.dis.mari.esmoke.race <- icc(model.dis.mari.esmoke.race)$ICC_adjusted
icc_total.dis.mari.esmoke.race <-
  sum(as.data.frame(VarCorr(model.dis.mari.esmoke.race),comp="Variance")$vcov[-3])
icc_intercept.dis.mari.esmoke.race <-
  as.data.frame(VarCorr(model.dis.mari.esmoke.race),
                comp="Variance")$vcov[1]/icc_total.dis.mari.esmoke
icc_wave.dis.mari.esmoke.race <-
  as.data.frame(VarCorr(model.dis.mari.esmoke.race),
                comp="Variance")$vcov[2]/icc_total.dis.mari.esmoke
icc_residual.dis.mari.esmoke.race <-
  as.data.frame(VarCorr(model.dis.mari.esmoke.race),
                comp="Variance")$vcov[4]/icc_total.dis.mari.esmoke


icc_table <- data.frame (
  Model = c("model.wave", "model.dis", "model.dis.mari", "model.dis.mari.esmoke",
            "model.dis.mari.esmoke.race"),
  icc_intercept = c(icc_intercept.b, icc_intercept.dis, icc_intercept.dismari,
                    icc_intercept.dis.mari.esmoke, icc_intercept.dis.mari.esmoke.race),
  icc_wave = c(icc_wave.b, icc_wave.dis, icc_wave.dismari, 
               icc_wave.dis.mari.esmoke, icc_wave.dis.mari.esmoke.race),
  icc_residual = c(icc_residual.b, icc_residual.dis, icc_residual.dismari, 
                   icc_residual.dis.mari.esmoke, icc_residual.dis.mari.esmoke.race),
  icc_total_adjusted = c(icc.b, icc.dis, icc.dis.mari, 
                         icc.dis.mari.esmoke, icc.dis.mari.esmoke.race)
)

print(icc_table)
```

\newpage
# Summary Table 1
\begin{table}[ht]
  \centering
  \small
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{lccccc}
  \toprule
  \multirow{2}{*}{\textbf{Parameter}} & \textbf{Model 1} & \textbf{Model 2} & \textbf{Model 3} & \textbf{Model 4} & \textbf{Model 5 (Final)} \\
   & (Unconditional growth) & (+ Discrimination) & (+ Marijuana) & (+ Ever smoking) & (+ Race) \\
  \midrule
  \textbf{Fixed Effects} \\
  Intercept & 3.572 (0.066) & 2.990 (0.117)\textbf{***} & 2.930 (0.118)\textbf{***} & 2.858 (0.119)\textbf{***} & 2.890 (0.128)\textbf{***} \\
  Wave & -0.026 (0.026) & -0.020 (0.026) & -0.033 (0.026) & -0.032 (0.026) & -0.033 (0.026) \\
  Discrimination & - & 0.224 (0.038)\textbf{***} & 0.212 (0.038)\textbf{***} & 0.207 (0.038)\textbf{***} & 0.205 (0.038)\textbf{***} \\
  Marijuana use & - & - & 0.2227 (0.079)\textbf{**} & 0.154 (0.081)\textbf{.} & 0.157 (0.081)\textbf{.} \\
  Ever Smoking & - & - & - & 0.337 (0.093)\textbf{***} & 0.338 (0.093)\textbf{***} \\
  Race (ref: White) & - & - & - & - & - \\
  \hspace{5mm} Black, AA, or Negro & - & - & - & - & -0.020 (0.104) \\
  \hspace{5mm} Other & - & - & - & - & -0.733 (0.310)\textbf{*} \\
  \midrule
  \textbf{Random Effects} \\
  Residual (Level 1) & 1.314 (1.146) & 1.317 (1.147) & 1.315 (1.147) & 1.311 (1.145) & 1.311 (1.145) \\
  Intercept (Level 2) & 1.129 (1.063) & 1.023 (1.011) & 1.003 (1.001) & 0.962 (0.981) & 0.986 (0.972) \\
  Wave (Level 2) & 0.054 (0.233) & 0.046 (0.214) & 0.045 (0.211) & 0.048 (0.219) & 0.048 (0.219) \\
  \midrule
  \textbf{Model Fit} \\
  AIC & 6547 & 6514 & 6508 & 6497 & \textbf{6496} \\
  BIC & 6580 & 6553 & 6552 & \textbf{6547} & 6557 \\
  Log-Likelihood & -3267 & -3250 & -3246 & -3240 & \textbf{-3237} \\
  Deviance & 6535 & 6500 & 6492 & 6479 & \textbf{6474} \\
  \bottomrule
  \end{tabular}
  }
  \label{tab:model_summary}
  \end{table}
  
  \noindent\footnotesize
  AA = African-American. \quad
  \textbf{***} $p < .001$, \textbf{**} $p < .01$, \textbf{*} $p < .05$, \textbf{.} $p < .1$.
  



# Summary Table ICC
\begin{table}[ht]
    \centering
    \begin{tabular}{|l|c|c|c|c|}
    \hline
    \textbf{Model} & \textbf{ICC Intercept} & \textbf{ICC Wave} & \textbf{ICC Residual} & \textbf{ICC Total Adjusted} \\
    \hline
    Model 1 (Unconditional growth) & 0.452 & 0.0217 & 0.526 & 0.447 \\
    Model 2 (+ Discrimination) & 0.429 & 0.0193 & 0.552 & 0.427 \\
    Model 3 (+ Marijuana) & 0.424 & 0.0188 & 0.557 & 0.424 \\
    Model 4 (+ Ever Smoke) & 0.414 & 0.0207 & 0.565 & 0.417 \\
    \textbf{Model 5 (Final: + Race)} & \textbf{0.408} & \textbf{0.0207} & \textbf{0.565} & \textbf{0.413} \\
    \hline
    \end{tabular}
    \label{tab:icc_comparison}
    \end{table}
  


```{r}
# model + interactions
model.inter1 <- lmer(mental_worry ~ wave + discrimination*wave + 
                       marijuana_bi + smoke_ever_clean + 
                       (wave|PID)+ race_clean_labeled,
                       data = df, REML = FALSE)
summary(model.inter1)
icc(model.inter1)

model.inter2 <- lmer(mental_worry ~ wave + discrimination + 
                       marijuana_bi*wave + smoke_ever_clean + (wave|PID) + 
                       race_clean_labeled,
                       data = df, REML = FALSE)
summary(model.inter2)
icc(model.inter2)

model.inter3 <- lmer(mental_worry ~ wave + discrimination + marijuana_bi + 
                       smoke_ever_clean*wave + (wave|PID)+ race_clean_labeled,
                       data = df, REML = FALSE)
summary(model.inter3)
icc(model.inter3)
```

# Discussion

## Factor wave for discussion

```{r}
df_mod <- df %>% 
  mutate(fac_wave = factor(wave))

model.wave.f <- lmer(mental_worry ~ fac_wave + (wave|PID),
                     data = df_mod, REML = FALSE)
summary(model.wave.f)

model.dis.f <- lmer(mental_worry ~ fac_wave + discrimination + (wave|PID),
                    data = df_mod, REML = FALSE)
summary(model.dis.f)

# + mari_bi 
model.dis.mari.f <- lmer(mental_worry ~ fac_wave + discrimination + 
                           marijuana_bi + (wave|PID),
                       data = df_mod, REML = FALSE)
summary(model.dis.mari.f)
icc(model.dis.mari.f)

# + ever smoke
model.dis.mari.esmoke.f <- lmer(mental_worry ~ fac_wave + discrimination + 
                                  marijuana_bi + smoke_ever_clean + (wave|PID),
                       data = df_mod, REML = FALSE)
summary(model.dis.mari.esmoke.f)
icc(model.dis.mari.esmoke.f)

# + race
model.dis.mari.esmoke.race.f <- lmer(mental_worry ~ fac_wave + 
                                       discrimination + marijuana_bi + 
                                       smoke_ever_clean + (wave|PID) + 
                                       race_clean_labeled,
                       data = df_mod, REML = FALSE)
summary(model.dis.mari.esmoke.race.f)
icc(model.dis.mari.esmoke.race.f)
```

## Use correlation - can't rule out other possibilities

```{r}

df_mod_base <- df_mod %>% 
  filter(wave == 0)
df_mod_1 <- df_mod %>% 
  filter(wave == 1)
df_mod_2 <- df_mod %>% 
  filter(wave == 2)
df_mod_3 <- df_mod %>% 
  filter(wave == 3)
 
df_mentalworry <- data.frame(mw_base = df_mod_base$mental_worry,
                             mw_1 = df_mod_1$mental_worry,
                             mw_2 = df_mod_2$mental_worry,
                             mw_3 = df_mod_3$mental_worry)

cor(df_mentalworry)
```
